A Study in Pairwise Testing Heuristics for Granger-causal Network Estimation

### 2019-07-26 paper_appl/paper_appl.tex
Developing a substantive application.

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

@@

\title{Graph Topological Aspects of Granger Causal Network Learning }
\author{R. J. Kinnear \\
  \small\href{mailto:ryan@kinnear.ca}{ryan@kinnear.ca} \\
  \small\url{https://github.com/RJTK} \and R. R. Mazumdar \\
  \small\href{mailto:mazum@uwaterloo.ca}{mazum@uwaterloo.ca} }

\begin{document}
\maketitle
\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO algorithms) when structural assumptions are met.  Finally,
  we provide an example application where we accurately discriminate
  between subjects in an EEG study based purely on the
  Granger-causality graphs inferred from data by our algorithms,
  demonstrating that meaningful features are captured by the
  Granger-causal graph topology.}

\keywords{causality graph, EEG, gene regulatory networks,
  Granger-causality, LASSO, network learning, time series, vector
  autoregression}

\paragraph{Acknowledgement}

We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC), [funding reference number 518418-2018].  Cette recherche a été financée par le Conseil de recherches en sciences naturelles et en génie du Canada (CRSNG), [numéro de référence 518418-2018].

@@

\section{Introduction and Review}
\label{sec:introduction}
% Consider a collection of stochastic processes producing observations
% at discrete time intervals.  Are the underlying processes dependent?
% Can we quantify any of the underlying relationships?  Can the arrow of
% time help us to distinguish a directionality or flow of dependence
% among our observed series?  In this paper we contribute to the
% understanding of the notion of Granger-Causality
% \cite{granger1969investigating} \cite{Granger1980329} as a tool for
% answering these questions.

% Heuristically, we state that if an event '$A$' occuring strictly prior
% in time to an even '$B$' and $A$ provides us with exclusive (that is,
% not available anywhere else) information about $B$, then $A$ must have
% had a causal impact on $B$.  This is an entirely model-free notion of
% cause, and instead leverages the intuition that a cause must precede
% it's effect.

In this paper we study the notion of Granger-causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly, we
will infer it's presence as a latent structure among our observed time
series data.  This notion is leveraged in a variety of applications
e.g. in Neuroscience as a means of recovering interactions amongst
brain regions \cite{bressler2011wiener}, \cite{anna_paper2008},
\cite{david2008identifying}; in the study of the dependence and
connectedness of financial institutions \cite{NBERw16223}; gene
expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.

The LASSO algorithm is known to suffer from a large proportion of
false-positives (e.g. see \cite{fdr_lasso}), except when certain
stringent conditions are met.  One of the key improvements to the
original LASSO algorithm is the adaptive (i.e. weighted) ``adaLASSO''
\cite{adaptive_lasso_zou2006}, for which oracle results
(i.e. asymptotic support recovery) are established under less
restrictive conditions.  Our experimental comparisons in Section
\ref{sec:empirical_evaluation} are against the
adaLASSO.

\subsection{Contributions}
In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph, which to our knowledge remains largely
unexplored.  The focus of this paper is to shed light on some of these
topological questions, in particular, we study a particularly simple
notion of causality graph topology which we term ``strongly causal''
and show that stationary times series whose underlying causality graph
has this structure satisfy natural intuitive notions of ``information
flow'' through the graph.  Moreover, we show that such graphs are
perfectly recoverable with only \textit{pairwise} Granger-causality
tests, which would otherwise suffer from serious confounding problems.
Our finite sample results are based on simulations, where we show
promising results for these graph topologies where our algorithm
performs substantially better in our simulation setup than do
competing LASSO algorithms, even for graphs that do not exactly
satisfy our strongly-causal topology assumptions.

In the case of gene expression networks, we show examples from the
literature which suggest our concept of a ``strongly causal graph''
topology may have application in this field (see Section
\ref{sec:strongly_causal_graphs}).

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).
Secondly, we propose in section \ref{sec:structure_learning} a graph
search heuristic which implements our theoretical results to finite
data samples, specifying and summarizing appropriate methods for
hypothesis testing (section \ref{sec:pairwise_hypothesis_testing}),
model order selection (section \ref{sec:model_order_selection}),
computationally efficient estimation (section
\ref{sec:efficient_model_estimation}), and error rate controls
(section \ref{sec:error_rate_control}).  Our heuristics are compared
against the adaLASSO algorithm in Section
\ref{sec:empirical_evaluation}.  We stress the scalability of our
algorithm which is capable of comfortably handling hundreds or
thousands of nodes on a single machine, as opposed to standard LASSO
algorithms which do not take advantage of the special structure
associated with stationary time series data.  In section
\ref{sec:application} we develop an example application where a
classifier is constructed to accurately discriminate between subjects
in an EEG study based only on the Granger-causality graph topology
inferred by our algorithms.  Concluding remarks on further open
problems and extentions are provided in Section \ref{sec:conclusion}.

\section{Theory}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.  We
suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has absolutely
summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$ with an absolutely
continuous spectral density $S(\omega)$:

\begin{equation}
  \label{eqn:fourier_pair}
  \begin{aligned}
    R(\tau) &= \frac{1}{2\pi}\int_{-\pi}^\pi S(\omega) e^{j\tau\omega}\d \omega,\\
    S(\omega) &= \sum_{\tau=-\infty}^\infty R(\tau)e^{-j\tau\omega}.
  \end{aligned}
\end{equation}

% and that the spectra is bounded uniformly away from 0:

% \begin{equation*}
%   \int_{-\pi}^\pi \ln\det S(\omega) \d\omega > -\infty.
% \end{equation*}

We will also work frequently with the spaces spanned by the values of
such a process

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    % \H^x &= \cl \{\sum_{\tau = -T}^T a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, T \in \N\} \subseteq L_2(\Omega),\\
    \H_t^x &= \cl \{\sum_{\tau = 0}^p a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, p \in \N\} \subseteq L_2(\Omega)\\
    H_t^x &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analogous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but particular components as
$\H_t^{-j}$.

As a consequence of the Wold decomposition theorem \cite{lindquist},
every WSS sequence has the moving average $MA(\infty)$
representation

\begin{equation}
\label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) v(t - \tau),
\end{equation}

where $c(t)$ is a purely deterministic sequence\footnote{the purely
  deterministic sequence $c(t)$ is one which lies in the remote past
  $\bigcap_{\tau=1}^\infty \H_{t - \tau}^x$ of the process.  For such
  processes a single sample $c(t_0)$ is enough to determine $c(t)$ for
  every $t$.  For example,
  $c(t) = \text{sin}(2\pi t + \Theta);\; \Theta \sim \mathcal{U}[-\pi,
  \pi]$}, $v(t)$ is an uncorrelated sequence and $A(0) = I$.  We will
assume that $c(t) = 0$, which in practice is to say that the process
has been detrended.  We additionally require that this representation
can be inverted to yield the $\VAR(\infty)$ form

\begin{equation}
  \label{eqn:ar_representation}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t).
\end{equation}

The equations (\ref{eqn:wold}), (\ref{eqn:ar_representation}) can be
represented as $x(t) = \A(z)v(t) = \B(z)x(t) + v(t)$ via the action
(convolution) of the operators (LTI filters)
$\A(z) \defeq \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$ and
$\B(z) \defeq \sum_{\tau = 1}^\infty B(\tau)z^{-\tau}$ where the
operator $z^{-1}$ is the back shift operator acting on
$\ell_2^n(\Omega, \mathcal{F}, \mathbb{P})$, that is:

\begin{equation}
  \label{eqn:filter_action}
  \B_{ij}(z)x_j(t) \defeq \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau).
\end{equation}

Finally, since $||z^{-1}|| = 1$ we have the inversion formula

\begin{equation}
  \label{eqn:lsi_inversion}
  \A(z) = (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
\end{equation}

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal matrix,
which is referred to as a lack of instantaneous feedback in $x(t)$.
We formally state our setup to conclude this section:

\begin{definition}[Basic Setup]
  \label{def:basic_setup}
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $\VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal covariance matrix.  The $MA(\infty)$
  representation of equation \eqref{eqn:wold} has $c(t) = 0$ and
  $A(0) = I$.
\end{definition}

\subsection{Granger Causality}

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ satisfying the assumptions of Definition
  \ref{def:basic_setup} we will say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces $\H_{t - 1}$,
  $\H^{-j}_{t - 1}$

\begin{equation}
  \linEerr{x_i(t)}{\H_{t - 1}} < \linEerr{x_i(t)}{\H^{-j}_{t - 1}},
\end{equation}

where $\xi[x \ |\ \H] = \E (x - \linE{x}{\H})^2$ is the mean squared
estimation error and $\linE{x}{\H} = \text{proj}_{\H}(x)$ denotes the
(unique) projection onto the Hilbert space $\H$.
\end{definition}

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x$'' is important in that GC relations can
change when components are added to or removed from our collection $x$
of observations, e.g. new GC relations can arise if we remove the
observations of a common cause, and existing GC relations can
disappear if we observe a new mediating series.

The notion is closely related to the information theoretic measure of
transfer entropy, indeed, if the distribution of $v(t)$ is known to be
Gaussian then they are equivalent \cite{barnett2009granger}.

We require two technical facts we refer to in the sequel, the notion
of conditional orthogonality is used throughout.

\begin{lemma}[\cite{lindquist} Proposition 2.4.2]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$,
  $\mathcal{B}$, $\mathcal{X}$.  The following statements are
  equivalent

  \begin{enumerate}
    \item{$\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$}
    \item{$\linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}$.}
    % \item{$\linE{\beta}{\mathcal{A}} = \linE{\linE{\beta}{\mathcal{X}}}{\mathcal{A}}\ \forall \beta \in \mathcal{B}}$}
    \end{enumerate}

    Where $\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$ denotes
    conditional orthogonality:

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}
\end{lemma}
% \begin{proof}
%   Proceding from the definition of conditional orthogonality we have

%   \begin{align*}
%     &\inner{\alpha - \linE{\alpha}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%     &{\iff} \inner{\alpha + x - \linE{\alpha + x}{\mathcal{X}}}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
%     &\overset{(a)}{\iff} \inner{\alpha + x}{\beta - \linE{\beta}{\mathcal{X}}} = 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}, x \in \mathcal{X}\\
%     &\overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A} \vee \mathcal{X}} = 0\ \forall \beta \in \mathcal{B}\\
%     &\overset{(c)}{\iff} \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}},
%   \end{align*}

%   where $(a)$ follows since $(\beta - \linE{\beta}{\mathcal{X}}) \perp \mathcal{X}$ and $\linE{\alpha + x}{\mathcal{X}} \in \mathcal{X}$, $(b)$ since $z \perp \mathcal{X} \vee \mathcal{A} \iff \linE{z}{\mathcal{X} \vee \mathcal{A}} = 0$, and $(d)$ since $\linE{z}{\H_2} = \linE{\linE{z}{\H_2}}{\H_1}$ for $\H_2 \subseteq \H_1$.

%   % This is property (vi) in Lindquist
%   % \begin{align*}
%   %   \langle \alpha - \linE{\alpha}{\mathcal{X}}, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0 \ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%   %   \overset{(a)}{\iff} \langle \alpha, \beta - \linE{\beta}{\mathcal{X}} \rangle &= 0\ \forall \alpha \in \mathcal{A}\ \beta \in \mathcal{B}\\
%   %   \overset{(b)}{\iff} \linE{\beta - \linE{\beta}{\mathcal{X}}}{\mathcal{A}} &= 0\ \forall \beta \in \mathcal{B}
%   % \end{align*}

%   % where $(a)$ follows because $\linE{\alpha}{\mathcal{X}} \in \mathcal{X}$ and $\beta - \linE{\beta}{\mathcal{X}} \in \mathcal{X}$, and $(b)$ follows since $\beta \perp \mathcal{A} \iff \linE{\beta}{\mathcal{A}} = 0$.
% \end{proof}
\begin{lemma}
  \label{lem:subspace_sum_projection}
  Let $x \in \H$ where $\H$ is a separable Hilbert space having inner product
  $\inner{\cdot}{\cdot}$ and let $\{\H_n\}_{n = 0}^N$ be a
  collection of subspaces of $\H$ such that $x \perp \H_0$.  Then

  \begin{equation*}
    \linE{x}{\bigvee_{n = 0}^N\H_n} = \linE{x}{\bigvee_{n = 1}^N\H_n},
  \end{equation*}

  where
  $\H_1 \vee \H_2 = \cl \{\alpha + \beta\ |\ \alpha \in \H_1, \beta
  \in \H_2 \}$ is the closed sum of subspaces.
\end{lemma}
\begin{proof}
  Apply the Gram-Schmidt process and directly calculate the projections.
\end{proof}

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and spectral density uniformly bounded above and below.
  Denote by $\xi_{ij} \defeq \linEerr{x_i(t)}{\H^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\H_t}$ for distinct $i, j$.  Then,
  the following are equivalent:

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$F_{ij} \defeq \big(\frac{\xi_i}{\xi_{ij}} - 1\big) = 0$}
    \item{$\H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j} \iff H_t^{i} \perp \H_{t - 1}^{j}\ |\ \H_{t - 1}^{-j}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{-j}} = \linE{x_i(t)}{\H_{t - 1}}$}
    % \item{\hl{Wold $A(\tau)$ condition}}
    % \item{Tie in the 0s of $S(\omega)^{-1}$, though this isn't immediately directed.}
  \end{enumerate}
\end{theorem}

\begin{proof}
  % Geweke uses the log of the ratio of the determinants of the residual variances
  % 
  % The equivalence $(1) \iff (3)$ is essentially a restatement of the
  % definition, but is in line with the seminal work of Geweke
  % \cite{geweke1982measurement}, \cite{geweke1984}.

  $(1) \iff (3)$ is a tautology, and $(3) \iff (5)$ is immediate from the
  definitions i.e. the projections are equal if and only if the errors
  are equal.  We have $(4) \iff (5)$ as a result of Lemma
  \ref{lem:conditional_orthogonality_equivalence}

% , the additional
%   equivalence in $(4)$ is apparent since $\H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$.

  $(1) \Rightarrow (2)$: The projection for $x_i(t)$ onto $\H_{t - 1}$ is
  (by definition) given by a form similar to equation
  \ref{eqn:ar_representation}:

\[
  \E|x_i(t) - \linE{x_i(t)}{\H_{t - 1}}|^2 = \E|v_i(t) + \sum_{\tau = 1}^\infty \sum_{k = 1}^n (B_{i, k}(\tau) - \hat{B}_{i, k}(\tau))x_k(t - \tau)|^2.
\]

Since $v(t)$ in equation \ref{eqn:ar_representation} is temporally
uncorrelated, it follows that the optimal projection is given by the
model coefficients themselves.  This holds similarly for the
projection onto $\H_{t - 1}^{-j}$.  Then by $(1)$ we have

\[
  \E|x_i(t) - \sum_{\tau = 1}^\infty \sum_{k = 1}^n B_{i, k}(\tau)x_k(t - \tau)|^2 = \E|x_i(t) - \sum_{\tau = 1}^\infty\sum_{k \ne j}B_{i, k}(\tau)x_k(t - \tau)|^2
\]

By the uniqueness of the projection we must have
$\forall \tau\ B_{i, j}(\tau) = 0$.

$(2) \Rightarrow (4)$: In computing $(y - \linE{y}{\H_{t - 1}^{-j}})$ for
$y \in H_t^i$ it is sufficient to consider $y = x_i(t)$ by linearity, then since
$H_{t - 1}^i \subseteq \H_{t - 1}^{-j}$ we have
$(x_i(t) - \linE{x_i(t)}{\H_{t = 1}^{-j}}) = v_i(t)$ since $\B_{ij}(z) = 0$.  $(4)$ then follows
since $v_i(t) \perp \H_{t - 1}$ and
$\forall z \in \H_{t - 1}^j\ (z - \linE{z}{\H_{t - 1}^{-j}}) \in \H_{t
  - 1}$
\end{proof}

% The following propositions justify various modifications of $x(t)$
% applied in practice to massage $x(t)$ into a form amenable to more
% standard tools.

% \begin{theorem}[General Invariance]
%   \hl{Under what conditions is this actually true?}

%   Let $\zeta(t)$ be a stationary discrete time stochastic process.  Let
%   $F, G$ be (possibly nonlinear and time varying) invertible filtering
%   operations and $f(t)$, $g(t)$ be perfectly predictable and such that
%   $x_j(t) \defeq F(\zeta_j - f)(t), x_i(t) \defeq G(\zeta_i - g)(t)$ are W.S.S.  Then,

% \begin{equation}
%     x_j \gc x_i \iff \zeta_j \te \zeta_i
%   \end{equation}
% \end{theorem}

% \begin{proposition}[Invariance Under Invertible and Deterministic Modifications]
%   Let $x(t)$ be a WSS process with absolutely summable covariance
%   sequence, and spectral density uniformly bounded above and below.  Let
%   $F(z), G$ represent univariate and invertible
%   linear-time-invariant filters.  And, let $f(t), g(t)$ be perfectly
%   predictable processes.  Then,

%   \begin{equation}
%     x_j \gc x_i \iff F(x_j - f)(t) \gc G(x_i - g)(t)
%   \end{equation}

% \begin{proof}
%   \hl{TODO}
% \end{proof}
% \end{proposition}

\subsection{Granger Causality Graphs}
We first need to establish some graph theoretic notation and
terminology, collected formally in definitions for the reader's
convenient reference.

\begin{definition}[Graph Theory Review]
  A \textit{graph} $\gcg = (V, \gcge)$ is simply a
  tuple of sets respectively called \textit{nodes} and \textit{edges}.
  Throughout this paper, we have in all cases
  $V = [n] \defeq \{1, 2, \ldots, n\}$.  We will also focus solely on
  \textit{directed} graphs, where the edges
  $\gcge \subseteq V \times V$ are \textit{ordered} pairs.

  A (directed) \textit{path} (of length $r$) from node $i$ to node
  $j$, denoted $\gcgpath{i}{j}$, is a sequence
  $a_0, a_1, \ldots, a_{r - 1}, a_r$ with $a_0 = i$ and $a_r = j$ such
  that $\forall\ 0 \le k \le r\ (a_k, a_{k + 1}) \in \gcge$, and where
  $(a_k, a_{k - 1})$ are \textit{distinct} for $0 \le k < r$.

  A \textit{cycle} is a path of length $2$ or more between a node and
  itself.  An edge between a node and itself $(i, i) \in \gcge$ (which
  is not a cycle) is referred to as a \textit{loop}.

  A graph $\gcg$ is a \textit{directed acyclic graph} (DAG) if it is a
  directed graph and does not contain any cycles.
\end{definition}

\begin{definition}[Parents, Grandparents, Ancestors]
  A node $j$ is a \textit{parent} of node $i$ if $(j, i) \in \gcge$.
  The set of all $i$'s parents will be denoted $\pa{i}$, and we
  explicitly exclude loops as a special case, that is,
  $i \not\in \pa{i}$ even if $(i, i) \in \gcge$.

>>>>

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage[backend=bibtex,sorting=none]{biblatex}
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{multirow}
\bibliography{\string~/Documents/academics/global_academics/global_bib.bib}
% \usepackage{apxproof}

@@

\title{A Study in Pairwise Testing Heuristics for Granger-causal Network Estimation}
\author{R. J. Kinnear \\
  \small\href{mailto:ryan@kinnear.ca}{ryan@kinnear.ca} \\
  \small\url{https://github.com/RJTK} \and R. R. Mazumdar \\
  \small\href{mailto:mazum@uwaterloo.ca}{mazum@uwaterloo.ca}}

\begin{document}
\maketitle
\abstract{Following recent theoretical developments of
  Granger-causality graph estimation for ``strongly causal'' causality
  graphs, we develop in detail a computationally and statistically
  efficient heuristic for Granger-causality estimation which is based
  purely on pairwise causality testing.  We perform detailed
  comparisons against state of the art algorithms in the LASSO family
  through a simulation study, and demonstrate the capacity for scaling
  to thousands of nodes in a time series network.  We develop an
  application example for EEG data where we accurately discriminate
  between subjects based purely on the Granger-causality graphs
  inferred from data by our algorithms, demonstrating that meaningful
  features are being captured in the Granger-causalilt graph topology,
  as well as providing evidence that individuals exhibit meaningfully
  distinct EEG causality structures, which may be a result of
  independent scientific interest.}

\keywords{causality graph, EEG, Granger-causality, LASSO, network
  learning, time series, vector autoregression}

\paragraph{Acknowledgement}
We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC), [funding reference number 518418-2018].  Cette recherche a été financée par le Conseil de recherches en sciences naturelles et en génie du Canada (CRSNG), [numéro de référence 518418-2018].

@@

\section{Introduction and Review}
\label{sec:introduction}
Following recent theoretical developments \cite{my_GC_paper}
establishing that, under certain topological assumptions on the
underlying causality graph, it is possible to accurately recover the
true causality graph by way of pairwise causality tests alone.  That
is, there are algorithms which are immune to the effects of
confounders when applied to certain causality graph structures.  In
this paper we provide simulation evidence that, given structural
assumptions, such algorithms can dramatically outperform state of the
art LASSO-type estimators.  Moreover, even in cases where structural
assumptions do not in fact hold, using similar algorithms as pure
heuristics can still exhibit superior performance.  In addition to the
potential for gains in statistical power, that our algorithms make use
only of pairwise causality testing improves the scalability (both in
terms of the asymptotics, as well as through trivial parallelization)
of Granger-causality testing up to thousands of nodes.

Granger-causality is a notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.  Our experimental comparisons in Section
\ref{sec:empirical_evaluation} are against the
adaLASSO. % Success of the LASSO has lead to
% the famous ``bet on sparsity'' principle ``Use a procedure that does
% well in sparse problems, since no procedure does well in dense
% problems.''  \cite{tibshirani2015statistical}.

% \subsection{Contributions}
% In the context of time series data, sparsity assumptions remain
% important, but there is significant additional structure that may
% arise as a result of considering the topology of the underlying
% Granger-causality graph, which to our knowledge remains largely
% unexplored.  The focus of this paper is to shed light on some of these
% topological questions, in particular, we study a particularly simple
% notion of causality graph topology which we term ``strongly causal''
% and show that stationary times series whose underlying causality graph
% has this structure satisfy natural intuitive notions of ``information
% flow'' through the graph.  Moreover, we show that such graphs are
% perfectly recoverable with only \textit{pairwise} Granger-causality
% tests, which would otherwise suffer from serious confounding problems.
% Our finite sample results are based on simulations, where we show
% promising results for these graph topologies where our algorithm
% performs substantially better in our simulation setup than do
% competing LASSO algorithms, even for graphs that do not exactly
% satisfy our strongly-causal topology assumptions.

% In the case of gene expression networks, we show examples from the
% literature which suggest our concept of a ``strongly causal graph''
% topology may have application in this field (see Section
% \ref{sec:strongly_causal_graphs}).

% The principle contributions of this paper are as follows: firstly, in
% section \ref{sec:theory} we study \textit{pairwise} Granger-causality
% relations, providing novel theorems connecting the structure of the
% causality graph to the pairwise ``causality flow'' in the system, as
% well as an interpretation in terms of the graph topology of the
% sparsity pattern of matrices arising in the Wold decomposition,
% generalizing in some sense the notion of ``feedback-free'' processes
% studied by \cite{caines1975feedback} in close connection with
% Granger-causality.  We establish sufficient conditions (sections
% \ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
% which a fully conditional Granger-causality graph can be recovered
% from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).
% Secondly, we propose in section \ref{sec:structure_learning} a graph
% search heuristic which implements our theoretical results to finite
% data samples, specifying and summarizing appropriate methods for
% hypothesis testing (section \ref{sec:pairwise_hypothesis_testing}),
% model order selection (section \ref{sec:model_order_selection}),
% computationally efficient estimation (section
% \ref{sec:efficient_model_estimation}), and error rate controls
% (section \ref{sec:error_rate_control}).  Our heuristics are compared
% against the adaLASSO algorithm in Section
% \ref{sec:empirical_evaluation}.  We stress the scalability of our
% algorithm which is capable of comfortably handling hundreds or
% thousands of nodes on a single machine, as opposed to standard LASSO
% algorithms which do not take advantage of the special structure
% associated with stationary time series data.  In section
% \ref{sec:application} we develop an example application where a
% classifier is constructed to accurately discriminate between subjects
% in an EEG study based only on the Granger-causality graph topology
% inferred by our algorithms.  Concluding remarks on further open
% problems and extentions are provided in Section \ref{sec:conclusion}.

\section{Granger Causality}
\label{sec:theory}
For the purposes of this paper, we provide an expediant and simplified
definition of Granger-causality.  For a more detailed exposition see
e.g. \cite{my_GC_paper}, \cite{granger1969investigating},
\cite{geweke1982measurement}, \cite{lutkepohl2005new}.  Suppose that $x(t)$ is an
$n-$dimensional wide-sense stationary time series generated by the
following (stable) $VAR(\infty)$ model

\begin{equation}
  \label{eq:var_model}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t).
\end{equation}

\begin{definition}[Granger-causality]
  \label{def:granger_causality}
  We will say that the $j$th component of $x$, namely $x_j(t)$,
  Granger-causes component $x_i(t)$ (with respect to $x$) and write
  $j \gc i$ if $\exists \tau \in \Z_{++}$ such that $B_{ij}(\tau) \ne 0$.
\end{definition}

This definition captures the notion that $x_j$ provides information
about $x_i$ which is not available through any of the other components
of $x$.  The motivation being that, since an effect must follow it's
cause, that $x_j$ ``causes'' $x_i$.  This intuition can also be given
interpretation as the ``flow'' of ``information'' or ``energy''.

\begin{definition}[Granger-causality Graph]
  We define the Granger-causality graph $\gcg = ([n], \gcge)$ to be
  the directed graph formed on $n$ vertices where an edge
  $(j, i) \in \gcge$ if and only if $x_j$ Granger-causes $x_i$ (with
  respect to $x$).  That is,
  $$(j, i) \in \gcge \iff j \in \pa{i} \iff x_j \gc x_i.$$
\end{definition}

Referring back to the definition (Def. \ref{def:granger_causality}) of
Granger-causality it is important that the concept be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed,
completely explains the connection from $x_j$ to $x_i$.  We introduce
the pairwise definition of Granger-causality as a special case:
